---
title: "Data Preparation"
author: "Shir Goldfarb, Ofek Glik & Alon Samocha"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F, message=F)
```


## 1. Import and Transform Raw Data & Import R Libraries

#### 1.1. Import Relevant Libraries
```{r import_libraries, message=F, warning=F}
library(tidyverse)
library(zoo)
library(knitr)
library(ggthemes)
library(ggmap)
library(lubridate)
```

#### 1.2. Set global options and notebook's theme
```{r global_options}
my_theme <- theme_fivethirtyeight() + 
  theme(axis.title= element_text(face="bold", size=12),
        legend.text= element_text(),
        legend.title= element_text(),
        legend.direction = "vertical",
        legend.position="right")
theme_set(my_theme)
columns_to_show <- c("acq_date", "satellite", "daynight", "brightness", "frp", "confidence", "longitude", "latitude")
```

#### 1.3. Import Raw Data
```{r import_data}
aus_rainfall <- read.csv("./data/raw-data/aus_rainfall.csv")
aus_temperature <- read.csv("./data/raw-data/aus_temp.csv")
aus_fires1 <- read.csv("./data/raw-data/fire_archive_M-C61_270827.csv")
aus_fires2 <- read.csv("./data/raw-data/fire_nrt_M-C61_270827.csv")
isr_temperature <- read.csv("./data/raw-data/isr_temp.csv")
isr_rainfall <- read.csv("./data/raw-data/isr_rainfall.csv")
isr_fires1 <- read.csv("./data/raw-data/fire_archive_M-C61_271256.csv")
isr_fires2 <- read.csv("./data/raw-data/fire_nrt_M-C61_271256.csv")
isr_fires3 <- read.csv("./data/raw-data/fire_archive_SV-C2_271258.csv")
isr_fires4 <- read.csv("./data/raw-data/fire_nrt_J1V-C2_271257.csv")
isr_fires5 <- read.csv("./data/raw-data/fire_nrt_SV-C2_271258.csv")
```

#### 1.3. Combine Relevant Data Frames and Remove Unneccessary Columns
```{r combine_data}
aus_fires1 <- aus_fires1 %>% select(-c(instrument, type,version))
aus_fires2 <- aus_fires2 %>% select(-c(instrument, version))
aus_fires <- bind_rows(aus_fires1,aus_fires2)
aus_fires <- aus_fires %>% select(c(acq_date, acq_time, satellite, daynight, brightness,
                            bright_t31, confidence, scan, track, frp, longitude, latitude))

isr_fires1 <- isr_fires1 %>% select(-version)
isr_fires2 <- isr_fires2 %>% select(-version)
isr_fires3 <- isr_fires3 %>% select(-c(version, confidence, satellite))
isr_fires4 <- isr_fires4 %>% select(-c(version, confidence, satellite))
isr_fires5 <- isr_fires5 %>% select(-c(version, confidence, satellite))
isr_fires <- bind_rows(isr_fires1, isr_fires2, isr_fires3, isr_fires4, isr_fires5)
isr_fires <- isr_fires %>%
              select(c(acq_date, acq_time, satellite, instrument, daynight, brightness,
                            bright_t31, confidence, scan, track, frp, longitude, latitude))
```

#### 1.4. Transform Data Types and Sort by Date
```{r transform_data}
aus_fires <- aus_fires %>% transform(acq_date = as.Date(acq_date)) %>% arrange(acq_date, acq_time)
isr_fires <- isr_fires %>% transform(acq_date = as.Date(acq_date)) %>% arrange(acq_date, acq_time)
aus_temperature <- aus_temperature %>% transform(date = as.Date(date)) %>% arrange(date)
aus_rainfall <- aus_rainfall %>% transform(date = as.Date(date)) %>% arrange(date)
isr_temperature <- isr_temperature %>% transform(date = as.Date(date, format="%d-%m-%Y")) %>% arrange(date)
isr_rainfall <- isr_rainfall %>% transform(date = as.Date(date, format="%d-%m-%Y")) %>% arrange(date)
```

## 2. Exploratory Data Analysis
#### 2.1. Examine each of the datasets
```{r preview_aus_fires}
kable(head(aus_fires %>% select(columns_to_show)), align="c", caption="Australia Fires")
```

```{r preview_aus_temperature}
kable(tail(aus_temperature), align="c", caption="Australia Temperatures")
```

```{r preview_aus_rainfall}
kable(tail(aus_rainfall), align="c", caption="Australia Rainfall")
```

```{r preview_isr_fires}
kable(tail(isr_fires %>% select(columns_to_show)), align="c", caption="Israel Fires")
```
As we can see, the `confidence` column in the above table has missing values. Let's check how common that is.
```{r summary_isr_fires}
kable(summary(isr_fires %>% select(columns_to_show)))
```
Using the summary function we can learn that out of 19,000~ records in this table, more than 17,000 have NA `confidence` columns. Since `confidence` gives us certainty about the occurrence of fires, we will need to understand how we can fill those values using other features.


```{r preview_isr_temperature}
kable(tail(isr_temperature), align="c", caption="Israel Temperature")
```


```{r preview_isr_rainfall}
kable(tail(isr_rainfall), align="c", caption="Israel Rainfall")
```


#### 2.2. Visualizations


##### 2.2.1 Australia Fires - Plot latest fires locations
```{r, aus_map, message=F, warning=F}
late_fires <- aus_fires %>% filter(acq_date >= as.Date('2020-01-01'), acq_date < as.Date('2021-01-01')) 
sbbox <- make_bbox(lon = c(115, 155), lat = c(-12.5, -42), f = .1)
ausbg = get_map(location=sbbox, zoom=4,
                source = "osm",
                color = "color",
                maptype="satellite")
ausbg = ggmap(ausbg)
ausbg +
  stat_density2d(data = late_fires, aes(x = longitude, y= latitude, 
                     fill = ..level..,  alpha = I(.2)),
                 size = 1, bins = 5, geom = "polygon") +
  geom_point(data = late_fires, mapping = aes(x=longitude, y= latitude), 
             color="red", alpha = .2, size=late_fires$brightness/500) +
  # define color of density polygons
  scale_fill_gradient(low = "grey50", high = "grey20") +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        panel.background = element_rect(fill = "aliceblue",
                                colour = "aliceblue"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        # supress legend
        legend.position = "none")
```

##### 2.2.2 Israel Fires - Plot latest fires locations
```{r, isr_map, echo=F, message=F, warning=F}
late_fires <- isr_fires %>% filter(acq_date >= as.Date('2020-01-01'), acq_date < as.Date('2021-01-01')) 
sbbox <- make_bbox(lon = c(34.069, 36.073), lat = c(33.459, 29.382))
isr_bg = get_map(location=sbbox, zoom=8,
                source = "osm",
                color = "color",
                maptype="satellite")
isr_bg = ggmap(isr_bg)
isr_bg +
  stat_density2d(data = late_fires, aes(x = longitude, y= latitude, 
                     fill = ..level..,  alpha = I(.2)),
                 size = 1, bins = 5, geom = "polygon") +
  geom_point(data = late_fires, mapping = aes(x=longitude, y= latitude), 
             color="red", alpha = .2, size=late_fires$brightness/500) +
  # define color of density polygons
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        panel.background = element_rect(fill = "aliceblue",
                                colour = "aliceblue"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        # supress legend
        legend.position = "none")
```

##### 2.2.3 Australia Fires - `Confidence` vs `Brightness` by `daynight`
```{r}
aus_fires %>% ggplot(aes(x=brightness, y=confidence, color=daynight)) +
              geom_point(alpha=0.7) + 
              geom_smooth(method="lm") + 
              ylim(0, 150) +
              labs(title = "Australia Fires",
                   subtitle = "Confidence vs Brightness by Day/Night",
                   x = "Brightness",
                   y = "Confidence",
                   caption = "Source: NASA Firms") + 
              scale_color_discrete(name="Day/Night", labels=c("Day", "Night"))
```

##### 2.2.4 Australia Fires - `Confidence` vs `Fire Radiative Power`
```{r}
aus_fires %>% ggplot(aes(x=frp, y=confidence, color=daynight)) +
              geom_point(alpha=0.7) + 
              geom_smooth(method="lm") +
              ylim(0, 150) +
              labs(title = "Australia Fires",
                   subtitle = "Confidence vs Fire Radiative Power by Day/Night",
                   x = "Fire Radiative Power",
                   y = "Confidence",
                   caption = "Source: NASA Firms") + 
              scale_color_discrete(name="Day/Night", labels=c("Day", "Night"))
```

As we can see in the 2 plots above, we have certain `confidence` values that act as anomalies. We will filter those values in order to reduce the noise and make our model more reliable in the future.
Also, we can see that there are significant differences in the way that `brightness` behaves in day and night. In order to make our data more stable and predictable, we will work with fires that occurred in the day only from now on.
Most importantly, we can learn that there is a clear correlation between between `confidence`, `brightness` and `frp`. We will use this relationship in order to fill the missing `confidence` values in the Israeli Fires table.


##### 2.2.5. Filter Australia Fires
```{r filter_data_1}
filtered_aus_fires <- aus_fires %>%
  filter(confidence != 0, confidence!= 100, confidence!=87,confidence!=94, daynight == "D")
```

##### 2.2.6. Review the data after filtering
```{r}
aus_fires %>% ggplot(aes(x=frp, y=confidence)) +
              geom_point(alpha=0.7) + 
              geom_smooth(method="lm") +
              ylim(0, 150) +
              labs(title = "Australia Fires",
                   subtitle = "Confidence vs Fire Radiative Power",
                   x = "Fire Radiative Power",
                   y = "Confidence",
                   caption = "Source: NASA Firms")
```

##### 2.2.7 Israel Fires - `Confidence` vs `Brightness` by `daynight`
```{r}
isr_fires %>% ggplot(aes(x=brightness, y=confidence, color=daynight)) +
              geom_point(alpha=0.7) + 
              geom_smooth(method="lm") + 
              ylim(0, 150) +
              xlim(300, 450)
              labs(title = "Israel Fires",
                   subtitle = "Confidence vs Brightness by Day/Night",
                   x = "Brightness",
                   y = "Confidence",
                   caption = "Source: NASA Firms") + 
              scale_color_discrete(name="Day/Night", labels=c("Day", "Night"))
```
Again, we can learn that there are different behaviors between day and night and so we will also filter the Israeli Fires dataset accordingly.
Also, we can see that the `brightness` values in Israel are distributed a bit differently than Australia. We will normalize this column in both datasets in order to bride the differences.


#### 2.2.8. Normalize `Brightness` column and add binary `Confidence Level` column

```{r filter_data_2}
normalit<-function(m){
   (m - min(m))/(max(m)-min(m))
}

avg <- mean(filtered_aus_fires$confidence)
std <- sd(filtered_aus_fires$confidence)

filtered_aus_fires <- filtered_aus_fires %>%
  filter(confidence <= (avg - std) | confidence >= (avg + std)) %>%
  mutate(confidence_level = cut(confidence, breaks=c('-inf', avg-std, 'inf'), labels=c(0,1)),
         brightness=normalit(brightness))

filtered_isr_fires <- isr_fires %>% 
  filter(daynight == "D")%>%
  mutate(confidence_level = cut(confidence, breaks=c('-inf', avg-std, 'inf'), labels=c(0,1)),
          brightness = normalit(brightness))
```

#### 2.2.9. Maximum and Minimum temperatures in Australia by year
```{r, echo=F, message=F}
year_min_max_temp <- aus_temperature %>% mutate(year=year(date)) %>%
  group_by(year, min.max) %>%
  transmute(min_temp = min(temperature,na.rm=TRUE),
            max_temp = max(temperature, na.rm=TRUE)) %>%
  distinct()
temp_plot <- ggplot(year_min_max_temp, aes(x = year, y = max_temp, color=min.max))+
  geom_point()+
  geom_smooth(method = 'loess', se=F)+
  scale_x_continuous(breaks=seq(1860, 2022, 20)) +
  labs(
    title = "Maximum and Minimum temperature observed per year",
    x = "Year",
    y = "Temperature (C)",
    caption = "Source: Australian Bureau of Meterology"
  ) +
  scale_color_discrete(name="Min/Max", labels=c("Max","Min"))
temp_plot

```
#### 2.2.10. Accumulated Rainfall in Israel by year
```{r, echo=F}
isr_annual_rainfall <-
  isr_rainfall %>%
  mutate(year=year(date)) %>%
  filter(year >= 2010, year < 2022) %>%
  group_by(year) %>%
  transmute(total_amount = sum(daily_amount, na.rm=T)) %>%
  distinct()
ggplot(isr_annual_rainfall, aes(x=year,y=total_amount)) + 
    scale_x_continuous(breaks=seq(2000, 2022, 5)) +
  geom_bar(stat = "identity", fill="lightblue") +
  geom_smooth(se=F)
``` 

## 3. Join Spatial Data
Find the closest weather station and the distance for each fire record and insert as new columns

#### 3.1. Create function that returns a Distance Matrix between each fire location and each weather station
```{r distance_matrix}
myEuclid <- function(points1, points2) {
    distanceMatrix <- matrix(NA, nrow=dim(points1)[1], ncol=dim(points2)[1])
    for(i in 1:nrow(points2)) {
        for (j in c(1:dim(t(points1))[2])) {
            
        distanceMatrix[j,i] <- sqrt(rowSums(t(t(points1)[,j]-t(points2[i,]))^2))
            }
    }
    distanceMatrix
}
```

#### 3.2. Calculate the distance of each fire in Australia to the nearest rain station
```{r aus_rain_stations}
# aus_rain_stations <- aus_rainfall[!duplicated(aus_rainfall$station_name),] %>% select(station_name ,long, lat)
# aus_dist_rain <- myEuclid(filtered_aus_fires %>% select(longitude, latitude), aus_rain_stations %>% select(long, lat))
# aus_dist_rain <- as.data.frame(aus_dist_rain)
# aus_dist_rain$min_index <- apply(aus_dist_rain, 1, which.min)
# aus_dist_rain <- aus_dist_rain %>% rowwise() %>% mutate(min_value = min(V1, V2, V3, V4, V5, V6, V7),
#                                                         closest_station = case_when(
#                                                           min_index == 1 ~ "Darwin Botanic Gardens",
#                                                           min_index == 2 ~ "North Adelade",
#                                                           min_index == 3 ~ "Melbourne Botanical Gardens",
#                                                           min_index == 4 ~ "Subiaco Treatment Plant",
#                                                           min_index == 5 ~ "Brisbane Botanic Gardens",
#                                                           min_index == 6 ~ "Canberra Airport",
#                                                           min_index == 7 ~ "Sydney Observatory Hill"))
aus_dist_rain <- read.csv("./data/raw-data/aus_rainfall_dist_matrix.csv")
filtered_aus_fires$closest_rain_station <- aus_dist_rain$closest_station
filtered_aus_fires$distance_from_rain_station <- aus_dist_rain$min_value
```

#### 3.3. Calculate the distance of each fire in Australia to the nearest temperature station
```{r aus_temp_stations}
# aus_temp_stations <- aus_temperature[!duplicated(aus_temperature$station_name),] %>% select(station_name ,long, lat)
# aus_dist_temp <- myEuclid(filtered_aus_fires %>% select(longitude, latitude), aus_temp_stations %>% select(long, lat))
# aus_dist_temp <- as.data.frame(aus_dist_temp)
# aus_dist_temp$min_index <- apply(aus_dist_temp, 1, which.min)
# aus_dist_temp <- aus_dist_temp %>% rowwise() %>% mutate(min_value = min(V1, V2, V3, V4, V5, V6, V7),
#                                                         closest_station = case_when(
#                                                           min_index == 1 ~ "Sydney Observatory Hill",
#                                                           min_index == 2 ~ "Darwin Airport",
#                                                           min_index == 3 ~ "Adelaide Airport",
#                                                           min_index == 4 ~ "Swanbourne",
#                                                           min_index == 5 ~ "Brisbane",
#                                                           min_index == 6 ~ "Canberra Airport",
#                                                           min_index == 7 ~ "Melbourne Olympic Park"))
aus_dist_temp <- read.csv("./data/raw-data/aus_temp_dist_matrix.csv")
filtered_aus_fires$closest_temp_station <- aus_dist_temp$closest_station
filtered_aus_fires$distance_from_temp_station <- aus_dist_temp$min_value
```


#### 3.4. Calculate the distance of each fire in Israel to the nearest rain station
```{r isr_rain_stations}
isr_rain_stations <- isr_rainfall[!duplicated(isr_rainfall$station_name),] %>% select(station_name ,long, lat)
isr_dist_rain <- myEuclid(filtered_isr_fires %>% select(longitude, latitude), isr_rain_stations %>% select(long, lat))
isr_dist_rain <- as.data.frame(isr_dist_rain)
isr_dist_rain$min_index <- apply(isr_dist_rain, 1, which.min)
isr_dist_rain <- isr_dist_rain %>% rowwise() %>% mutate(min_value = min(V1, V2, V3, V4, V5, V6),
                                                        closest_station = case_when(
                                                          min_index == 1 ~ "Kfar Blum",
                                                          min_index == 2 ~ "Haifa-Technion",
                                                          min_index == 3 ~ "Beit Dagan",
                                                          min_index == 4 ~ "Beer Sheva",
                                                          min_index == 5 ~ "Eilat",
                                                          min_index == 6 ~ "Jerusalem Center"))
filtered_isr_fires$closest_rain_station <- isr_dist_rain$closest_station
filtered_isr_fires$distance_from_rain_station <- isr_dist_rain$min_value
```


#### 3.5. Calculate the distance of each fire in Israel to the nearest temperature station
```{r isr_temp_stations}
isr_temp_stations <- isr_temperature[!duplicated(isr_temperature$station_name),] %>% select(station_name ,long, lat)
isr_dist_temp <- myEuclid(filtered_isr_fires %>% select(longitude, latitude), isr_temp_stations %>% select(long, lat))
isr_dist_temp <- as.data.frame(isr_dist_temp)
isr_dist_temp$min_index <- apply(isr_dist_temp, 1, which.min)
isr_dist_temp <- isr_dist_temp %>% rowwise() %>% mutate(min_value = min(V1, V2, V3, V4, V5, V6, V7, V8),
                                                        closest_station = case_when(
                                                          min_index == 1 ~ "Beer Sheva",
                                                          min_index == 2 ~ "Eilat",
                                                          min_index == 3 ~ "Haifa-Technion",
                                                          min_index == 4 ~ "Jerusalem Center",
                                                          min_index == 5 ~ "Kfar Blum",
                                                          min_index == 6 ~ "Tel Aviv Beach",
                                                          min_index == 7 ~ "The Negev Junction",
                                                          min_index == 8 ~ "Beit Dagan"))
filtered_isr_fires$closest_temp_station <- isr_dist_temp$closest_station
filtered_isr_fires$distance_from_temp_station <- isr_dist_temp$min_value
```

#### 3.6. Add Temperature and Rainfall (7-day rolling sum) columns to each fire, measured by the closest weather station.
Since we have records of fires starting at 2010, we will filter the rainfall and temperature data to match that (Australia only).
```{r filter_2010}
aus_rainfall_2010 <- aus_rainfall %>% filter(date >= "2010-01-01")
aus_temp_2010 <- aus_temperature %>% filter(date >= "2010-01-01", min.max == "max") %>%
                                select(date, station_name, temperature) 
```

#### 3.7. Calculate the 7-day rolling sum of rainfall in Australia and join with the fire data
```{r aus_rainfall_sum}
aus_rainfall_2010 <- aus_rainfall_2010 %>%
                      replace(is.na(.), 0) %>%
                      arrange(date, station_name) %>%
                      group_by(station_name) %>%
                      mutate(rainfall_7d_sum = zoo::rollsum(amount, k = 7, fill=0)) %>%
                      ungroup() %>%
                      select(date, station_name, rainfall_7d_sum)
filtered_aus_fires <- filtered_aus_fires %>% left_join(aus_rainfall_2010, by=c("closest_rain_station" = "station_name", "acq_date"="date"))
```

#### 3.8. Add a temperature column to each fire in Australia, measured by the closest temperature station
```{r aus_temperature}
filtered_aus_fires <- filtered_aus_fires %>% left_join(aus_temp_2010, by=c("closest_temp_station" = "station_name", "acq_date"="date")) %>% replace(is.na(.), 0)
```

#### 3.9. Calculate the 7-day rolling sum of rainfall in Israel and join with the fire data
```{r isr_rainfall_sum}
isr_rainfall_join <- isr_rainfall %>%
                      drop_na() %>%
                      arrange(date, station_name) %>%
                      group_by(station_name) %>%
                      mutate(rainfall_7d_sum = zoo::rollsum(daily_amount, k = 7, fill=0)) %>%
                      ungroup() %>%
                      select(date, station_name, rainfall_7d_sum)
filtered_isr_fires <- filtered_isr_fires %>% left_join(isr_rainfall_join, by=c("closest_rain_station" = "station_name", "acq_date"="date"))
```

#### 3.10. Add a temperature column to each fire in Israel, measured by the closest temperature station
```{r isr_temperature, message=F, warning=F}
isr_temp_join <- rename(isr_temperature,temperature = max_temp)
isr_temp_join <- isr_temp_join %>% mutate_at("temperature", as.numeric) %>% drop_na(temperature) %>% select(date, station_name, temperature)
filtered_isr_fires <- filtered_isr_fires %>% left_join(isr_temp_join, by=c("closest_temp_station" = "station_name", "acq_date"="date")) %>% mutate(rainfall_7d_sum = replace_na(rainfall_7d_sum, 0)) %>% drop_na(temperature)
```

## 4. Export the Transformed Data to .csv Files
Later to be used by the models
```{r export_data}
write.csv(filtered_aus_fires, "./data/brightness-frp-model/filtered_aus_fires.csv")
write.csv(filtered_isr_fires, "./data/brightness-frp-model/filtered_isr_fires.csv")
```
